Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-30 WholePlotTrajectories.Rmd
details
Bubs correction

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   
    Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
    The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the controversy questions the extent of communities' resilience regarding their functional and taxonomic facets.

    To disentangle the ecological processes driving community response to disturbance, we analysed the diversity trajectories over 30 years following a disturbance gradient.
    Specifically, we examined community functional and taxonomic trajectories with regards to diversity, composition and redundancy. Functional trajectories were drawn based on 7 leaf, stem and life-history traits.
    We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities.
    The IDH did predict communities taxonomic diversity response while functional diversity was enhanced whenever the disturbance intensity.
    Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbance.

@@

In tropical forests, ecological communities are regularly re-shaped by natural disturbance events changing both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions such as competition among species [@Chesson2000;@Herault2018]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to disturbance [@Chazdon2003a]. 
For now, this has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016] that are rapid and convenient to measure. 
These structural parameters have been sucessfully modeled, giving important insights into the recovery of ecosystem processes and services [@Herault2018]. 
However the response of forests diversity and composition remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014; @Liang2016].
In the short-term, moderate disturbance may lead to positive impacts on communities diversity, an idea formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity when disturbance intensity is not too high [@Molino2001;@Kariuki2006a].

Validations of the IDH though remain scarce in the long-term and mainly rely on the analysis of taxonomic richness [@Molino2001]. Taxonomic richness alone, though, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. More ecological-meaningful analysis would couple richness with (i) evenness that would reveal underlying ecological processes and (ii) composition that is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. Furthermore, a functional approach accounting for species biological attributes would directly link communities diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was sucessfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually a complete overview of communities response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyperdiverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
Specifically, we (i) elucidated community taxonomic and functional recovery and the determinism of underlying processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) questioned community recovery time.

@@

The taxonomic differences among communities, marked before disturbance by the distinct starting points on the NMDS axis 2, were maintained throughout recovery trajectories. 
More than commonly thought, post-disturbance trajectories depended on community initial composition, that partly determined the pool of recruited species and constrained the trajectories towards the initial composition. 
The high resilience of communities taxonomy revealed that species not belonging to the pre-disturbance community were hardly recruited because of the commonness of dispersal limitation among tropical tree species [@Svenning2005].

Conversely, disturbed communities followed functional trajectories similar in terms of functional composition.
As pre-disturbance surviving trees mirror the initial community [@Herault2018], changes in functional composition relied upon the recruitment of species or functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches of high availability of light, space and nutrient vacated by the disturbance.
The recruitment of pioneers changed community functional composition in the same way for all disturbance intensity towards more resource-acquisitive strategies, moving community functional composition right along the first axis in Figure \@ref(fig:NMDSplans) [@Westoby1998; @Wright2004; @Reich2014].
Thereafter long-lived, more resistant and shade-tolerant species excluded the first established pioneers and started the recovery of pre-disturbance functional composition, moving similarly community functional composition left along the first axis and upward along the second axis in Figure \@ref(fig:NMDSplans).

@@

The IDH predicted well the disturbance impact on community taxonomic richness, enhanced until an intensity threshold (20-25% AGB loss), and to some extent on taxonomic evenness, somewhat decoupled from the disturbance intensity as already observed in the Guiana Shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees and those recruited afterward.
The pool of true pioneer species specifically recruited after disturbance is restricted in the Guiana Shield to a few common genera (e.g. Cecropia spp., Vismia spp.) [@Guitet2018].
Below the intensity threshold the size of the surviving community maintained the pre-disturbance high taxonomic richness while the recruitment of pioneers, infrequent or absent before disturbance, increased both community taxonomic richness and evenness.
Beyond the intensity threshold, the disturbance decreased the taxonomic richness of surviving trees which was not offset by the enrichment of pioneers, so that the overall community taxonomic richness decreased according to the disturbance intensity [@Molino2001].
For community taxonomic evenness the disturbance impact was similar but slighter, as the evenness is less sensitive to the loss of rare species.
Taxonomic evenness rather represented the increasing dominance of pionneers that balanced the usual hyper-dominance of tropical forest below the intensity threshold, thus increasing community overall evenness up to the intensity threshold beyond which pioneers became in turn highly dominant and decreased the overall evenness [@Baraloto2012a].

@@

For 15 years the species loss during disturbance, determined by the disturbance intensity, commensurately decreased the functional redundancy within the pre-disturbance functional space.
The redundancy decrease was not compensated in the first place as the first recruited pionneers were functionally different from the pre-disturbance functional composition.
Progressively though, first established species were replaced by more competitive long-lived pionneers or late-successional species resembling more the pre-disturbance functional composition and restoring the functional redundancy.
This replacement was stochastic and followed the lottery recruitment rules, implying a recruitment eased for the first recruited species but then increasingly hampered by the emergence of interspecific competition [@Busing2002]. 
Along time the recovery of infrequent species was increasingly difficult, so that the time for the full recovery of the functional redundancy, in some communities just initiated after 30 years, was difficult to estimate [@Elmqvist2003;@Diaz2005].

@@

Our study revealed community recovery through the combination of deterministic processes driving their convergence in functional composition, and dispersal limitation maintaining their divergence in taxonomic composition.
The IDH was validated for community taxonomic richness and, to some extent, taxonomic evenness but disproved regarding community functional richness and evennes that were enhanced for any disturbance intensity by the high functional differences of pioneers compared to late-successional functoinal composition.
The IDH was translated in time by the recruitment, beyond an intensity threshold, of short-lived pionneers that prevented in the first times the establishment of more diverse long-lived pionneers, recruited otherwise below the intensity threshold.
The resilience of tropical forests proved consistent although several decades-long. 
Still, the disturbance impact on communities redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Gourlet-Fleury2005].

>>>>

---
title: "Post-Disturbance Tree Community Trajectories in a Neotropical Forest"
author:
  - name: Ariane MIRABEL
    authsuperscript: 1*
  - name: Eric Marcon
    authsuperscript: 1
  - name: Bruno Hérault
    authsuperscript: 2 3
affiliation:
  - affsuperscript: 1
    dptuniv: UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles, Université de Guyane.
    address: >
      Campus Agronomique,
      97310 Kourou,
      France.
  - affsuperscript: 2
    dptuniv: Cirad, Univ montpellier, UR Forests & Societies.
    address: >
      Montpellier,
      France.
  - affsuperscript: 3
    dptuniv: INPHB, Institut National Polytechnique Félix Houphouet-Boigny
    address: >
      Yamoussoukro,
      Ivory Coast.
corrauthor:
    email: ariane.mirabel@ecofog.gf
    url: http://www.ecofog.gf/spip.php?article47
abstract: >   
    Understanding the ecological rules underlying the maintenance of tropical forests biodiversity, structure, functioning and dynamics is urgent to anticipate their fate in the global change context.
    The huge diversity of tropical forests is often assumed to be regularly reshaped by natural disturbance yielding a diversity peak at intermediate intensity. This intermediate disturbance hypothesis (IDH), though, remains debated and the controversy questions the extent of communities' resilience regarding their taxonomic and functional facets.

    To disentangle the ecological processes driving community response to disturbance, we analysed the tree community trajectories over 30 years following a disturbance gradient in a Neotropical forest.
    Specifically, we examined community functional and taxonomic trajectories with regards to diversity, composition and redundancy. Functional trajectories were drawn based on 7 leaf, stem and life-history traits.
    We highlighted the cyclic recovery of community taxonomic and functional composition. While pre-disturbance taxonomic differences were maintained over time, functional composition trajectories were quite similar among communities.
    The IDH did predict communities taxonomic diversity response while functional diversity was enhanced whatever the disturbance intensity.
    Although consistent, the recovery of community composition, diversity and redundancy remained unachieved after 30 years. This acknowledged the need of decades-long cycles with no disturbance to ensure a complete recovery, and questioned tropical forest community resilience after repeated disturbances.

@@

In tropical forests, ecological communities are regularly re-shaped by natural disturbance events changing both the abiotic environment, through the fluxes of light, heat and water [@Goulamoussene2017], and the biotic interactions such as competition among species [@Chesson2000]. 
One of the cornerstone of tropical forest ecology is to understand the processes and drivers of ecosystems response to disturbance [@Chazdon2003a]. 
For now, this has been largely studied through forest structural parameters such as aboveground biomass, tree height or stem density [@Piponiot2016;@Rutishauser2016] that are rapid and convenient to measure. 
These structural parameters have been sucessfully modeled, giving important insights into the recovery of ecosystem processes and services [@Herault2018]. 
However the response of forests diversity and composition remains unclear, albeit it determines the productivity, stability and functioning of ecosystems [@Tilman2014; @Liang2016].
In the short-term, moderate disturbance may lead to positive impacts on communities diversity, an idea formalized by the intermediate disturbance hypothesis (IDH) stating a maximized species diversity when disturbance intensity is not too high [@Molino2001;@Kariuki2006a].

Validations of the IDH though remain scarce in the long-term and mainly rely on the analysis of taxonomic richness [@Molino2001]. Taxonomic richness alone, however, gives limited or misleading information on forests recovery and functioning [@Chaudhary2016]. More ecological-meaningful analysis would couple richness with (i) evenness, that would reveal the changes in the species abundance distribution and thus the underlying ecological processes, and (ii) composition that is crucial for conservation issues [@Lavorel2002;@Bellwood2006]. Furthermore, a functional approach accounting for species biological attributes would directly link communities diversity, composition and redundancy to ecosystem functioning and to its environmental constraints [@Violle2007b;@Baraloto2012a].
In that respect, the functional trait-based approach that focus on major traits related to species ecology and mediate species performance in a given environment was sucessfully adopted [@Diaz2005].
For instance, the functional approach revealed in tropical rainforests the deterministic processes entailing, after disturbance, a functional shift from a dominance of "conservative" slow-growing species dealing with scarce resources to "acquisitive" fast-growing species with rapid and efficient use of abundant resources [@Reich2014;@Herault2011].
This shift is translated into the trajectories of key functional traits related to resource acquisition (leaf and stem traits) and life-history traits (seed mass, maximum size) [@Wright2004;@TerSteege2006;@Westoby2006a;@Chave2009b].
Eventually a complete overview of communities response to disturbance would encompass the changes in functional redundancy, that quantifies the amount of shared trait values among species [@Carmona2016].
The high functional redundancy of hyperdiverse tropical forests [@Bellwood2006] mitigates the impacts of species removal on ecosystem functioning and determines the resilience of communities after disturbance [@Elmqvist2003; @Diaz2005].

In this study, we monitored over 30 years the response of 75 ha of neotropical forest plots set up on a gradient of disturbance intensity, from 10 to 60% of ecosystem above-ground biomass (AGB) loss. 
We made use of a large functional traits database encompassing major leaf, stem and life-history traits in order to draw the taxonomic and functional trajectories in terms of richness, evenness, composition and redundancy. 
Specifically, we (i) elucidated community taxonomic and functional recovery and the underlying ecological processes, (ii) clarified the validity of the IDH in the long term for tropical forest and its translation into different trajectories in time, and (iii) questioned community recovery time.

@@

The taxonomic differences among local communities, marked before disturbance by the distinct starting points on the NMDS axis 2, were maintained throughout recovery trajectories. 
More than commonly thought, post-disturbance trajectories depended on community initial composition, that partly determined the pool of recruited species and constrained the trajectories towards the initial composition. 
The high resilience of communities taxonomy revealed that species not belonging to the pre-disturbance community were hardly recruited because of the commonness of dispersal limitation among tropical tree species [@Svenning2005].

Conversely, disturbed communities followed functional trajectories that are highly similar in terms of functional composition.
As pre-disturbance surviving trees mirror the initial community [@Herault2018], changes in functional composition relied upon the recruitment of species or functional types that were infrequent or absent before disturbance.
Competitive pioneers became dominant in filling the environmental niches of high availability of light, space and nutrients vacated by the disturbance.
The recruitment of pioneers changed community functional composition in the same way for all disturbance intensity towards more resource-acquisitive strategies, moving community functional composition right along the first axis in Figure \@ref(fig:NMDSplans) [@Westoby1998; @Wright2004; @Reich2014].
Thereafter long-lived, more resistant and shade-tolerant species excluded the first established pioneers and started the recovery of pre-disturbance functional composition, moving similarly community functional composition left along the first axis and upward along the second axis in Figure \@ref(fig:NMDSplans).

@@

The IDH well predicted well the disturbance impact on community taxonomic richness, enhanced until an intensity threshold (20-25% AGB loss), and to some extent on taxonomic evenness, somewhat decoupled from the disturbance intensity as already observed in the Guiana Shield [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998].
The disturbance intensity determined the balance in the community between pre-disturbance surviving trees and those recruited afterward.
The pool of true pioneer species specifically recruited after disturbance is restricted in the Guiana Shield to a few common genera (e.g. Cecropia spp., Vismia spp.) [@Guitet2018].
Below the intensity threshold the size of the surviving community maintained the pre-disturbance high taxonomic richness while the recruitment of pioneers, infrequent or absent before disturbance, increased both community taxonomic richness and evenness.
Beyond the intensity threshold, the disturbance decreased the taxonomic richness of surviving trees which was not offset by the enrichment of pioneers, so that the overall community taxonomic richness decreased according to the disturbance intensity [@Molino2001].
For community taxonomic evenness the disturbance impact was similar but slighter, as the evenness is less sensitive to the loss of rare species.
Taxonomic evenness rather represented the increasing dominance of pionneers that balanced the usual hyper-dominance of a few species in tropical forests below the intensity threshold, thus increasing community overall evenness up to the intensity threshold beyond which pioneers became in turn highly dominant and decreased the overall evenness [@Baraloto2012a].

@@

For 15 years the species loss during disturbance, determined by the disturbance intensity, commensurately decreased the functional redundancy within the pre-disturbance functional space.
The redundancy decrease was not compensated in the first place as the first recruited pionneers were functionally different from the pre-disturbance functional composition.
Progressively though, first established species were replaced by more competitive long-lived pionneers or late-successional species resembling more the pre-disturbance functional composition and restoring the functional redundancy.
This replacement was stochastic and followed the lottery recruitment rules, implying a recruitment eased for the first recruited species but then increasingly hampered by the emergence of interspecific competition [@Busing2002]. 
Along time the recovery of infrequent species was increasingly slow, so that the time for the full recovery of the functional redundancy, in some communities just initiated after 30 years, was extremely difficult to estimate [@Elmqvist2003;@Diaz2005].

@@

Our study revealed community recovery through the combination of deterministic processes driving their convergence in functional composition, and dispersal limitation maintaining their divergence in taxonomic composition.
The IDH was validated for community taxonomic richness and, to some extent, taxonomic evenness but disproved regarding community functional richness and evennes that were enhanced for any disturbance intensity by the high functional differences of pioneers compared to late-successional functoinal composition.
The IDH was translated in time by the recruitment, beyond an intensity threshold, of short-lived pionneers that prevented in the first times the establishment of more diverse long-lived pionneers, recruited otherwise below the intensity threshold.
The resilience of tropical forests proved consistent although several decades-long. 
Still, the disturbance impact on communities redundancy cautioned against the risks of infrequent species loss and the persistence of disturbance-specific communities [@Herault2018].
